---
title: "Asymmetric Smoothed-Association Matrices"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{Asymmetric Smoothed-Association Matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.alt    = paste(
    "Asymmetric smoothed-association matrix produced by",
    "janusplot(); diagonal cells hold variable labels,",
    "off-diagonal cells show the fitted mgcv::gam spline",
    "with a 95% confidence envelope, raw-data scatter,",
    "and per-cell annotations for n, EDF, and smooth",
    "significance glyph."
  ),
  fig.width  = 7,
  fig.height = 7,
  dpi        = 110,
  eval       = TRUE
)
library(ggplot2)
set.seed(2026L)
```

## Why Pearson is not enough

A Pearson correlation matrix gives one scalar per pair of variables.
Two numbers are discarded in that collapse:

1. The *shape* of the association — linear, monotone non-linear,
   U-shaped, or irregular.
2. The *direction* — whether $y$ is a smooth function of $x$
   differs in general from whether $x$ is a smooth function of $y$,
   because leverage and noise are directional.

`janusplot()` renders both recoveries visually for every pair in a
matrix layout, using proper `mgcv::gam()` fits (not loess) so EDF,
F-tests, and random effects are available.

## Quick start

```{r quickstart}
library(janusplot)

# Four numeric columns from mtcars (32 rows: small but illustrative)
janusplot(mtcars[, c("mpg", "hp", "wt", "qsec")])
```

Each off-diagonal cell shows:

- raw scatter (light grey),
- the fitted spline (blue line) and 95% CI ribbon,
- EDF (effective degrees of freedom) in the bottom-right,
- *n* used in the bottom-left,
- a signif-glyph in the top-right (`***` / `**` / `*` / `·`).

The cell fill is keyed to EDF: darker = more non-linear.

## Non-linear detection

A synthetic quadratic + sinusoidal example. The matrix makes it
obvious which variables are genuinely non-linearly related to which.

```{r nonlinear}
n <- 300
x1 <- runif(n, -3, 3)
x2 <- x1^2 + rnorm(n, sd = 0.6)   # quadratic on x1
x3 <- sin(x1) + rnorm(n, sd = 0.4) # sinusoidal on x1
x4 <- rnorm(n)                     # independent
d  <- data.frame(x1 = x1, x2 = x2, x3 = x3, x4 = x4)

janusplot(d)
```

EDF for `x2 ~ s(x1)` and `x3 ~ s(x1)` should clearly exceed 1; the
cell fills reflect that. Cells involving `x4` should be close to EDF =
1 (linear / flat).

## Asymmetry — a heteroscedastic example

When the noise scale depends on a predictor, the two directional smooths
diverge: $y \sim s(x)$ recovers the mean relationship; $x \sim s(y)$ is
distorted by the variance asymmetry.

```{r asym}
n <- 400
x <- runif(n, 0, 5)
y <- 0.5 * x + rnorm(n, sd = 0.3 + 0.4 * x)   # variance grows with x
d <- data.frame(x = x, y = y, z = rnorm(n))

janusplot(d)
```

The `A = ...` label per cell reports the asymmetry index
$A_{ij} = |EDF_{y|x} - EDF_{x|y}| / (EDF_{y|x} + EDF_{x|y}) \in [0, 1]$,
shown by default in the bottom-left corner alongside `EDF = ...`.

## Partial smooths (controlling for covariates)

Pass `adjust =` as a one-sided formula RHS to include fixed covariates
and/or random effects in every pairwise GAM.

```{r adjust, eval = requireNamespace("palmerpenguins", quietly = TRUE)}
library(palmerpenguins)
pp <- na.omit(penguins)

# Without covariate
janusplot(pp[, c("bill_length_mm", "bill_depth_mm",
                 "flipper_length_mm", "body_mass_g")])

# With species as a fixed effect — resolves Simpson's-paradox geometry
janusplot(pp, vars = c("bill_length_mm", "bill_depth_mm",
                       "flipper_length_mm", "body_mass_g"),
         adjust = ~ species)
```

## Changing the palette

The cell fill encodes the EDF (or deviance-explained) of the smooth and
is accompanied by a shared colourbar legend. Choose a palette with
`palette = `.

```{r palette-viridis, fig.height = 5, fig.width = 6}
d <- data.frame(
  x1 = runif(200, -3, 3),
  x2 = rnorm(200),
  x3 = rnorm(200)
)
d$x2 <- d$x1^2 + rnorm(200, sd = 0.8)  # non-linear

janusplot(d, palette = "viridis")  # default, colourblind-safe
```

```{r palette-brewer, fig.height = 5, fig.width = 6}
janusplot(d, palette = "RdYlBu")   # diverging, colourblind-safe
```

```{r palette-turbo, fig.height = 5, fig.width = 6}
janusplot(d, palette = "turbo")    # high-contrast, NOT colourblind-safe
```

**Colourblind-safe choices:**

- *Sequential:* `viridis` (default), `magma`, `inferno`, `plasma`,
  `cividis`, `mako`, `rocket`, `YlOrRd`, `YlGnBu`, `Blues`, `Greens`.
- *Diverging:* `RdYlBu`, `RdBu`, `PuOr`.

**High-contrast but not colourblind-safe:** `turbo`, `Spectral`.

## Handling missing data

```{r missing}
# airquality has genuine NAs in Ozone and Solar.R
janusplot(airquality[, c("Ozone", "Solar.R", "Wind", "Temp")],
          na_action = "pairwise")
```

`na_action = "pairwise"` uses all rows for which *that* pair is
complete; `"complete"` restricts to rows complete across every
variable (matching listwise deletion).

## Scaling up — `order = "hclust"`

For k large, reorder the axes by hierarchical clustering on
|correlation|:

```{r hclust, eval = requireNamespace("MASS", quietly = TRUE)}
data(Boston, package = "MASS")
janusplot(Boston[, c("medv", "lstat", "rm", "age",
                     "indus", "nox", "dis")],
          order = "hclust")
```

## Programmatic access — `janusplot_data()`

Returns raw GAM fits and per-cell metrics without constructing a
ggplot — useful for custom rendering or downstream analysis.

```{r data}
# Re-create the heteroscedastic example
n <- 400
het <- data.frame(
  x = runif(n, 0, 5),
  y = NA_real_
)
het$y <- 0.5 * het$x + rnorm(n, sd = 0.3 + 0.4 * het$x)

out <- janusplot_data(het, vars = c("x", "y"))
out$pairs[[1L]]$edf_yx
out$pairs[[1L]]$edf_xy
out$pairs[[1L]]$asymmetry_index
```

## Shape metrics explained

Every fitted smooth is summarised by two continuous indices and two
discrete counts. These drive the 24-category classifier and appear
as columns in `janusplot(..., with_data = TRUE)$data` and as fields
on each entry of `janusplot_data()$pairs`.

Let `f(x)` be the fitted smooth on a dense grid of 200 equally-spaced
points across the predictor range, with `f'` and `f''` the numerical
first and second derivatives. Let `w(x)` be the empirical density of
the predictor on the same grid, normalised to `sum(w) = 1`.

- **`monotonicity_index`** (paper symbol `M`):

    `M = sum(w * f') / sum(w * |f'|)  in  [-1, 1]`

    `+1` means strictly increasing, `-1` strictly decreasing, `0` a
    non-monotone curve (bowl, dome, wave).

- **`convexity_index`** (paper symbol `C`):

    `C = sum(w * f'') / sum(w * |f''|)  in  [-1, 1]`

    `+1` means globally convex (bowl-up), `-1` globally concave
    (bowl-down), `0` inflection-dominated (S-curve, sine, flat).

Both indices are density-weighted so they describe the smooth
*where the data actually live*, not extrapolated tails, and are
scale-invariant: replacing `y` with `a * y + b` leaves them unchanged.

- **`n_turning_points`** — count of interior extrema (sign changes
  of `f'`), robust to noise via lobe-mass weighting.
- **`n_inflections`** — count of interior curvature flips (sign
  changes of `f''`), same robust counting.

Together the pair `(n_turning_points, n_inflections)` drives the
primary `shape_category` dispatch; `(monotonicity_index,
convexity_index)` disambiguate within the monotone `(0, 0)` and
single-extremum `(1, 0)` cells. The full taxonomy with 2-letter
codes, archetypes, and thumbnail curves is available from
`janusplot_shape_hierarchy()` and is rendered as the standing legend
below every `janusplot()` call.

Tune the thresholds applied to these indices via
`janusplot_shape_cutoffs()`. See the
[shape-recognition-sensitivity](shape-recognition-sensitivity.html)
vignette for how faithfully the classifier recovers ground-truth
shapes across sample-size and noise regimes.

## Limitations

- Pairwise view, not conditional — always complement with a proper
  multivariate model.
- EDF depends on basis dimension `k`; defaults are sensible but
  domain-specific tuning is encouraged.
- The asymmetry index should not be interpreted causally without
  strong assumptions.
- `monotonicity_index` and `convexity_index` are scale-invariant in
  `y` but sensitive to the predictor-density weighting — they
  describe the smooth on the observed support of `x`, not outside
  it.

## Citation

```{r cite}
citation("janusplot")
```
