---
title: "Shape-recognition sensitivity study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shape-recognition sensitivity study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 8,
  fig.height = 6,
  dpi       = 110
)
```

`janusplot()` assigns every fitted smooth to one of 24 shape
categories via a `(n_turning_points, n_inflections)` dispatch with
additional `(monotonicity_index, convexity_index)` disambiguation
for the monotone cases (see the `janusplot` vignette for the full
definition of the indices). How reliably does this classifier
recover the ground-truth shape of a noisy sample? This vignette
answers the question with a
full-factorial sensitivity sweep.

## Design

For each combination of ground-truth shape, sample size `n`, and
noise level `sigma`, the sweep:

1. Generates `n` points from the noiseless canonical curve on
   `x ∈ [0, 1]`, with `y` normalised to `[0, 1]` so that `sigma` is
   the fraction of y-range that Gaussian noise contributes -- an
   SNR-comparable scale across shapes.
2. Fits `mgcv::gam(y ~ s(x), method = "REML")`.
3. Classifies the fit via `janusplot_shape_metrics()`.
4. Records correctness at the **fine** (24-category) and
   **archetype** (7-family) levels.

The design factors are orthogonal and replicated. See
`?janusplot_shape_sensitivity` for the function surface. The 14
canonical ground-truth shapes cover five of the seven archetypes
(`chaotic` and `degenerate` have no realistic deterministic
generator).

```{r setup-pkg}
library(janusplot)
library(ggplot2)

janusplot_shape_sensitivity_shapes()
```

## Pre-registered hypotheses

Four hypotheses were pinned before running the sweep:

- **H1.** At `n = 500`, `sigma = 0.05`, archetype accuracy
  exceeds 0.90 for every shape.
- **H2.** Fine-category accuracy exceeds 0.75 at `n = 500`,
  `sigma = 0.05` for monotone + unimodal shapes; wave and multimodal
  tolerate less noise.
- **H3.** Rippled variants require `n >= 200` and `sigma <= 0.10`
  to resolve.
- **H4.** At `sigma = 0.40`, archetype accuracy collapses below 0.50
  for all but the simplest shapes.

**Verdicts, evaluated against the precomputed demo below (`n = 500`,
`sigma = 0.05` for H1/H2; the demo's grid runs `sigma` up to 0.40,
which covers H4):**

- **H1 -- refuted.** `bimodal`, `concave_up`, `inverted_u` and
  `u_shape` all hit 1.00, `linear_up` reaches only 0.70, and `wave`
  scores 0.00 at every replicate. The `wave` failure is not a
  noise-tolerance property of the classifier: the shipped `wave`
  ground-truth generator (`.shape_sensitivity_generators()`) has one
  interior turning point on `x in [0, 1]`, one short of what the
  taxonomy's own `wave` definition requires, so every fit is
  systematically dispatched to the wrong archetype regardless of
  noise. This is a tracked defect in the generator, not evidence
  about recognition difficulty -- treat the `wave` row of every
  figure below as unreliable until the generator is fixed, and do
  not read the other five shapes' figures as "wave included and
  passing".
- **H2 -- partly supported, confounded by the same defect.** The
  four clean shapes clear 0.75 easily; `linear_up` (0.70) falls just
  short, which the sweep's design does correctly flag as a boundary
  case rather than a clear pass; `wave`'s 0.00 cannot speak to the
  hypothesis for the reason above.
- **H3 -- untestable with the shipped machinery.** No `rippled_*`
  generator exists among the canonical shapes exercised here, so
  the hypothesis has no data to be adjudicated against in this
  vignette.
- **H4 -- refuted, in the other direction.** At `sigma = 0.40`
  archetype accuracy stays at 0.60 or above for every shape except
  `wave` (still stuck at 0.00, for the generator reason above) --
  well clear of the predicted sub-0.50 collapse. Setting `wave`
  aside, the classifier is markedly more noise-robust than
  pre-registered.

## Precomputed demo

The package ships a small-footprint precomputed sweep -- 6 shapes
across 5 of the 7 archetypes (`u_shape` and `inverted_u` both roll
up to `unimodal`) × 3 sample sizes × 4 noise levels × 30 replicates
= 2160 fits -- so you can explore the API without running the full
sweep yourself.

```{r demo-data}
data("shape_sensitivity_demo")
str(shape_sensitivity_demo, vec.len = 2)
```

### Recovery curves (headline figure)

```{r recovery-curves, fig.cap = "Archetype-recovery accuracy against noise level (sigma) for each of the six ground-truth shapes, at each of the three sample sizes.", fig.alt = "Line plot of archetype-recovery accuracy versus sigma, one line per shape, faceted or coloured by sample size; five lines stay high across the noise range and one (wave) is flat at zero."}
janusplot_shape_sensitivity_plot(shape_sensitivity_demo,
                                 "recovery_curves")
```

Five of the six shapes are recovered well across the whole noise
range, including at `sigma = 0.40` (see H4 above): `bimodal`,
`concave_up` and `inverted_u` hold at 1.00 throughout at `n = 500`,
and `u_shape` only drops to 0.97 at the highest noise level.
`linear_up` is the outlier among the five -- it never exceeds 0.73
even at `sigma = 0.05`, so its curve is flat and mediocre rather than
a high-noise-tolerance decay. The sixth shape, `wave`, is flat at
0.00 across every `n` and `sigma` cell: this is the generator defect
described under H1, not a recovered-then-lost noise-tolerance curve,
and should not be read alongside the other five.

### Archetype confusion

```{r archetype-confusion, fig.width = 6, fig.height = 5, fig.cap = "Confusion matrix of true archetype (rows) against predicted archetype (columns), pooled across all n, sigma and replicates in the demo.", fig.alt = "Six-by-four archetype confusion matrix with predicted archetypes monotone_curved, monotone_linear, multimodal and unimodal on the columns; wave truth falls entirely in the unimodal column, and linear_up truth splits between monotone_linear and monotone_curved."}
janusplot_shape_sensitivity_plot(shape_sensitivity_demo,
                                 "confusion_archetype")
```

Two off-diagonal patterns dominate here. `wave` is predicted
`unimodal` on every single one of its 360 replicates -- the direct
consequence of the generator defect flagged under H1, since a curve
with only one interior turning point classifies as unimodal by
construction, regardless of noise. `linear_up`, by contrast, is a
genuine noise-driven confusion: it lands in `monotone_curved`
roughly 30% of the time, because sampling noise on a short curve
occasionally tips the fitted `monotonicity_index` below the
monotone-linear cutoff.

### Archetype-level accuracy grid

```{r accuracy-grid, fig.cap = "Heatmap of archetype-recovery accuracy across the full (n, sigma) design, one panel per shape.", fig.alt = "Grid of heatmap panels, one per ground-truth shape, with sample size on one axis and noise level on the other, cells shaded by archetype-recovery accuracy; five panels are uniformly dark (high accuracy) and one (wave) is uniformly pale (zero accuracy)."}
janusplot_shape_sensitivity_plot(shape_sensitivity_demo,
                                 "accuracy_grid")
```

Per-shape heatmap of `P(archetype correct)` across the `(n, sigma)`
design. Reading across a row shows the noise-tolerance profile of
one sample size; reading up a column shows the sample-size
sensitivity at one noise level. The `wave` panel is uniformly at
zero across the whole grid -- flat cells here mean the generator
defect above, not a genuinely noise-invariant failure mode.

### Numerical summary

```{r summary}
head(janusplot_shape_sensitivity_summary(shape_sensitivity_demo,
                                         level = "archetype"), 10)
```

## Running your own sweep

The demo is a starting point. For the publication-grade figure
use the full default grid (14 shapes × 4 sample sizes × 5 noise levels
× 200 reps = 56 000 fits):

```{r full-sweep, eval = FALSE}
# Configure parallel execution (optional) -- you control the plan.
future::plan(future::multisession, workers = 4L)

res <- janusplot_shape_sensitivity(parallel = TRUE)

# Save for your paper
saveRDS(res, "shape_sensitivity_full.rds")
janusplot_shape_sensitivity_plot(res, "recovery_curves")
```

### Custom shape subsets + cutoffs

Every argument is tunable. `mono_strong` and `curv_low` are only
read on the monotone dispatch path (`n_turning_points == 0` and
`n_inflections == 0`), so a worked example needs a monotone shape
family to actually exercise them -- `linear_up`, `concave_up` and
`convex_up` below, rather than the wave/multimodal family, where
these two cutoffs are never consulted. Tightening `mono_strong`
raises the bar for calling a curve "strictly monotone" (versus
`s_shape`/rippled), and lowering `curv_low` raises the bar for
calling it "near-linear" (versus curved).

```{r custom-subset, eval = FALSE}
strict <- janusplot_shape_cutoffs(mono_strong = 0.95, curv_low = 0.1)

res_strict <- janusplot_shape_sensitivity(
  shapes     = c("linear_up", "concave_up", "convex_up"),
  n_grid     = c(200L, 500L),
  sigma_grid = c(0.05, 0.10, 0.20),
  n_rep      = 100L,
  cutoffs    = strict
)

janusplot_shape_sensitivity_summary(res_strict, level = "fine")
```

## References

- Pya, N., & Wood, S. N. (2015). Shape constrained additive models.
  *Statistics and Computing*, 25(3), 543–559.
- Calabrese, E. J. (2008). Hormesis: why it is important to
  toxicology and toxicologists. *Environmental Toxicology and
  Chemistry*, 27(7), 1451–1474.
- Milnor, J. (1963). *Morse Theory*. Princeton University Press.
- Meyer, M. C. (2008). Inference using shape-restricted regression
  splines. *Annals of Applied Statistics*, 2(3), 1013–1033.

```{r session-info}
sessionInfo()
```
