---
title: "Shape-recognition sensitivity study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shape-recognition sensitivity study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width = 8,
  fig.height = 6,
  dpi       = 110
)
```

`janusplot()` assigns every fitted smooth to one of 24 shape
categories via a `(n_turning_points, n_inflections)` dispatch with
additional `(M, C)` disambiguation for the monotone cases. How
reliably does this classifier recover the ground-truth shape of a
noisy sample? This vignette answers the question with a
full-factorial sensitivity sweep.

## Design

For each combination of ground-truth shape, sample size `n`, and
noise level `sigma`, the sweep:

1. Generates `n` points from the noiseless canonical curve on
   `x ∈ [0, 1]`, with `y` normalised to `[0, 1]` so that `sigma` is
   the fraction of y-range that Gaussian noise contributes — an
   SNR-comparable scale across shapes.
2. Fits `mgcv::gam(y ~ s(x), method = "REML")`.
3. Classifies the fit via `janusplot_shape_metrics()`.
4. Records correctness at the **fine** (24-category) and
   **archetype** (7-family) levels.

The design factors are orthogonal and replicated. See
`?janusplot_shape_sensitivity` for the function surface. The 14
canonical ground-truth shapes cover five of the seven archetypes
(`chaotic` and `degenerate` have no realistic deterministic
generator).

```{r setup-pkg}
library(janusplot)
library(ggplot2)

janusplot_shape_sensitivity_shapes()
```

## Pre-registered hypotheses

The sweep's hypotheses are pinned in
`simulation/PLAN.md` (Scenario 4):

- **H1.** At `n = 500`, `sigma = 0.05`, archetype accuracy
  exceeds 0.90 for every shape.
- **H2.** Fine-category accuracy exceeds 0.75 at `n = 500`,
  `sigma = 0.05` for monotone + unimodal shapes; wave and multimodal
  tolerate less noise.
- **H3.** Rippled variants require `n ≥ 200` and `sigma ≤ 0.10`
  to resolve.
- **H4.** At `sigma = 0.40`, archetype accuracy collapses below 0.50
  for all but the simplest shapes.

## Precomputed demo

The package ships a small-footprint precomputed sweep — 6 shapes
(one per non-degenerate archetype) × 3 sample sizes × 4 noise levels ×
30 replicates = 2160 fits — so you can explore the API without
running the full sweep yourself.

```{r demo-data}
data("shape_sensitivity_demo")
str(shape_sensitivity_demo, vec.len = 2)
```

### Recovery curves (headline figure)

```{r recovery-curves}
janusplot_shape_sensitivity_plot(shape_sensitivity_demo,
                                 "recovery_curves")
```

Every shape is recovered near-perfectly at low noise; the
informative picture is where each shape's curve falls off as sigma
grows. The unimodal and monotone-curved families tolerate more noise
than the multimodal ones.

### Archetype confusion

```{r archetype-confusion, fig.width = 6, fig.height = 5}
janusplot_shape_sensitivity_plot(shape_sensitivity_demo,
                                 "confusion_archetype")
```

The off-diagonals reveal the classifier's failure modes. A `unimodal`
truth misclassified as `wave` or `multimodal` means the spline
invented extra turning points under noise.

### Archetype-level accuracy grid

```{r accuracy-grid}
janusplot_shape_sensitivity_plot(shape_sensitivity_demo,
                                 "accuracy_grid")
```

Per-shape heatmap of `P(archetype correct)` across the `(n, sigma)`
design. Reading across a row shows the noise-tolerance profile of
one sample size; reading up a column shows the sample-size
sensitivity at one noise level.

### Numerical summary

```{r summary}
head(janusplot_shape_sensitivity_summary(shape_sensitivity_demo,
                                         level = "archetype"), 10)
```

## Running your own sweep

The demo is a starting point. For the publication-grade figure
use the full default grid (14 shapes × 4 sample sizes × 5 noise levels
× 200 reps = 56 000 fits):

```{r full-sweep, eval = FALSE}
# Configure parallel execution (optional) — you control the plan.
future::plan(future::multisession, workers = 4L)

res <- janusplot_shape_sensitivity(parallel = TRUE)

# Save for your paper
saveRDS(res, "shape_sensitivity_full.rds")
janusplot_shape_sensitivity_plot(res, "recovery_curves")
```

### Custom shape subsets + cutoffs

Every argument is tunable. Below, we rerun only the bimodal/wave
family under stricter monotonicity thresholds to see whether
tightening `mono_strong` buys any fine-accuracy improvement for these
categories.

```{r custom-subset, eval = FALSE}
strict <- janusplot_shape_cutoffs(mono_strong = 0.95, curv_low = 0.1)

res_strict <- janusplot_shape_sensitivity(
  shapes     = c("wave", "bimodal", "bi_wave"),
  n_grid     = c(200L, 500L),
  sigma_grid = c(0.05, 0.10, 0.20),
  n_rep      = 100L,
  cutoffs    = strict
)

janusplot_shape_sensitivity_summary(res_strict, level = "fine")
```

## References

- Pya, N., & Wood, S. N. (2015). Shape constrained additive models.
  *Statistics and Computing*, 25(3), 543–559.
- Calabrese, E. J. (2008). Hormesis: why it is important to
  toxicology and toxicologists. *Environmental Toxicology and
  Chemistry*, 27(7), 1451–1474.
- Milnor, J. (1963). *Morse Theory*. Princeton University Press.
- Meyer, M. C. (2008). Inference using shape-restricted regression
  splines. *Annals of Applied Statistics*, 2(3), 1013–1033.

```{r session-info}
sessionInfo()
```
